Deep Reinforcement Learning for Conservation Decisions

### 2021-04-30 manuscript/manuscript.Rmd
WIP
WIP outline

Deep Reinforcement Learning (RL), a machine learning technique in which a *software agent* is trained to select actions which best maximize some cumulative reward, could hold significant untapped promise for solving complex ecological and conservation decision-making problems from wildfire management to biodiversity preservation to zoonotic pandemics.
Unlike more widely recognized machine learning techniques for either *supervised* learning (such as image classification, [@species_ex], [@satellite_ex]) or unsupervised learning, RL focuses on solving *dynamic* problems which require planning ahead, often in uncertain and changing environments.
Of all machine learning techniques, RL most resembles a general artificial intelligence.
Notable successes using RL include AlphaZero, in which a software agent learned to beat the world's best human players in games of chess, Go, and Shogi [@alphazero], and major advances in robotics, including decision-making in autonomous vehicles [@robotics_ex].
These tasks all require the agent to not only anticipate future states of it's environment, but also respond to those changes to influence future outcomes.

<!-- Millie help on this para? -->

If conservation challenges are typically complex and dynamics, existing approaches are largely simple and static.
The establishment of protected areas (cite MPAs, 30x30, etc) focuses on policies fixed in space and time to address challenges that are fixed in neither.
The endangered species act ... Ecosystem services ....
Such simplicity is in no small part due to the challenge associated in determining a Some reference to [@TomDietrich] on potential for CS-based application.

In contrast to other ML methods, RL does not require massive data sets for training, relying instead on model simulations based on the best available ecological data and knowledge of underlying processes.
Despite the expansion in data collection from remote sensors, citizen science, and other efforts, ecological data remains sparse and unevenly sampled [CITE].
Moreover, no amount of data can train supervised or unsupervised ML approaches can predict a future which has no analog in the historical record.
Mechanistic, process-based models remain our most powerful tool to explore scenarios of dynamic ecosystems in changing climates.
Methods from *optimal control theory* have long been used in conservation and resource economics literature [@Clark1990] to determine optimal policies given such models [@Polasky2011].
However, control theory approaches are also tightly constrained by the "curse of dimensionality" [@Mangel1980; @Marescot2013], limiting their application to only simple, stylized models.
Deep reinforcement learning offers a way forward.
Deep RL agents are trained not with mountains of data, but by interacting with simulators.
Most recent advances in Deep RL have illustrated algorithms trained by playing arcade game simulators such as Atari BreakOut, or physics-based simulations of robotics movement [@TD3; @SAC; @PPO; @A2C].
Simulating from complex models is far simpler than solving them.
The richest ecosystem models, such as the global circulation models (GCMs) used to generate future climate scenarios [@something], or individual-based models used to predict forest growth [@something] or wildfire spread [@something], are massive simulations.
Given sufficient computational power, RL can be trained and evaluated across a broad range of possible realistic scenarios well beyond the reach of classical methods.
In this manner, RL would compliment, rather than replace, existing empirical and theoretical research which would continue to form the basis for those simulations.

If RL has great promise, the approach also imposes great risk.
Lacking the tractability of classical methods, an RL agent can appear to solve a wide class of scenarios with ease, but may fail unexpectedly under seemingly similar conditions [@henderson2017].
Despite recent progress, Deep RL is a young field and stability and robustness of its methods remain largely open problems
. RL approaches are no panacea, and improper design or testing can easily lead to negative outcomes
. Technical issues are not the only pitfall of AI applications to conservation problems
. Giving algorithms direct influence over conservation policy may also raise ethical and political issues, particularly if those algorithms (or components thereof) are treated as proprietary intellectual property
.

In this paper, we introduce the fundamentals of deep reinforcement learning and illustrate with simple examples how these approaches can be applied to important conservation decision-making problems, from over-harvesting in global fisheries to ecosystem tipping points to wildfire dynamics.
These examples span a range from problems for which an optimal analytic solution is already known, to problems not previously solved with classical methods, to problems not yet solved by RL.
We include a detailed appendix with complete examples in both python and R for training RL algorithms, as well as three stand-alone module "gyms" implementing the each of the conservation problems presented here using a standard benchmark [@gym], which we believe can facilitate future contributions both from researchers in RL searching for interesting and important open problems as well as ecologists who may seek to make those simulations more realistic and precise to specific application areas.
We hope this paper serves as a port of entry to researchers and students from either field seeking to engage in work on RL applications to conservation.
We also hope these examples serve as illustration of both the potential and perils of such applications.

@@

-   model (underlying ecological model)

@@

-   predict

@@

@@

# Sequential Decision Problems and Reinforcement Learning

<!-- meh, not sure we need this? -->

Reinforcement learning focuses primarily on sequential decision problems: tasks in which a decision-maker or "agent" must repeatedly decide what action to take in response to new information about the "state" of the system they seek to manage, in order to maximize some objective (referred to as the reward or utility).
Any conservation problem that considers the future possible scenarios is a sequential decision problem: even scenarios that face a one-off decision such as whether declare some region as a protected area, or list or de-list a species as endangered, are sequential decision problems if they must consider the timing of their actions.
In contrast, decisions that do not consider the future, such as deciding which of a collection of possible regions should be prioritized based only on the present: how many species currently inhabit a region, how much each region would cost to purchase today, are not sequential problems.
Central to any sequential decision problem is the need to consider possible future states, the need to forecast.
Because the agent's actions will influence that future, agents facing sequential decision problems must further be able to forecast how each possible action available to them will change that forecast.
Consequentially, our lack of knowledge about the underlying models and inability to predict future environments in our rapidly changing world pose a fundamental challenge to conservation decision-making.

This framework raises two conceptual challenges.
First, sequential decision-making problem is distinct from the problem of model estimation.
A sequential decision problem will typically treat the model as given, typically with some uncertainty around the process and possibly the measurmeents and model estimates as well.
Thus, methods for solving sequential decision-making should not be mistaken as alternatives to methods designed for model estimation, such as regression, Bayesian heirarchical modeling, generalized additive models (GAMs) etc.
Second, methods for describing and solving sequential decision-making problems are often given different terms by different communities to describe the same thing.
Sequential decision problems are common to behavioral ecology [@Mangel], conservation [@Marescot2013] natural resource economics [@Clark1990], and some engineering fields [] where they are known as "optimal control" problems, while computer science, neuroscience, and robotics typically refer to them as 'reinfocement learning'.

@@

There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
RL algorithms can generally be split into 2 groups, model-based and model-free RL.
In model-based RL, the agent is attempting to learn the state transition operator, $T$.
Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]
.
In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
.

@@

>>>>

<!-- 
- Global change problems: Important but tough decisions
- ML: exciting stuff, but only fitting data is driving through the rearview mirror
- RL is the ML for decisions (in dynamic/uncertain environments) 
- RL can use existing knowledge and models to improve decisions in complex/uncertain scenarios (read: global change biology)
- Here be dragons / map paragraph

# RL overview

<!-- Conceptual picture: (A) an "environment", and (B) an "agent"

What do we mean by an "environment"?  
  - Transition function
  - "reward"
  - examples

What do we mean by an "agent"?
  Components of an agent: 
    - *neural network*, whose *parameters* are estimated by *training* +
    - an RL algorithm, whose *hyperparameters* (including NN structure) are estimated by *tuning*
    - High-level overview of an RL algorithm components (model/model-free, on/off policy, stochastic/deterministic, etc).

  conceptual explanation of 'model-free', what it means to have a "policy" without modeling/predicting transitions

@@

@@

-   neural network

-   train

-   tune

-   parameters

-   hyper-parameters

-   model (underlying ecological model used to drive step() method in environment)

-   model (SB3 term  for "agent")

-   predict (SB3 term to return agent's action given an observation)

-   predict (ecological model which generates probabilistic forecast of future states)

@@

-   reward (utility, value)

-   policy function

-   value function

@@

@@

There is an ever increasing multitude of reinforcement learning algorithms, and while there are far too many algorithms to discuss in detail in this paper, we will herein classify and touch on some current methods.
RL algorithms can generally be split into 2 groups, model-based and model-free RL.
In model-based RL, the agent is attempting to learn the state transition operator, $T$. <!-- Sometimes the agent is just assuming this is given, right? -->
Once the transition model is known, the agent can then use the transition model to plan out what is the optimal action, often with Model Predictive Control [cite tassa 2012 MPC], or the agent can use the transition model to generate more samples to aid learning [cite Sutton dyna, alphago?]. <!-- For most recent methods focus on an approximate solution, since optimal is computationally impossible.  In contrast most readers will assume this step is trivial, if the transition matrix is known there's nothing more to do but "look" at the predicted transitions and make a choice. -->
In model-free RL, the agent does not attempt to learn a dynamics model and instead tries to either learn a parameterized policy, learn a value function, or try to learn both a parameterized policy and a value function.
The general trade-off between model-free and model-based approaches is that in practice, model-based algorithms are often more sample efficient than model-free algorithms but often fail to surpass model-free algorithms in overall agent performance [cite?]
.

@@

<!-- Can we add citations to each please -->

### 2021-04-30 manuscript/presubmission-letter.Rmd
WIP outline

To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games. 
_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, and _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
We believe that similar techniques could improve decision-making in managing ecological environments as well. <!-- Having trouble reconciling lines 21-23, trying to say that these simulators are like the grand challenges and given RL's track record of solving intractable decision-making problems, perhaps RL could shed light here? I think having a sentence asserting that RL would use ecological models not replace them would be good somewhere in here.-->
Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
Detailed, process-based models allow a window into the future scenarios of ecological management.
Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.
Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls.

>>>>

To date, the uncertain and changing environments considered by RL research have largely been drawn from examples in robotic movement and video games. 
_Nature_ recently featured RL agents achieving superhuman performance on classic Atari games, and _Science_ described the success of Deepmind's AlphaZero against grandmasters in Go, Chess, and Shogi.
We believe that similar techniques could improve decision-making in managing ecological environments as well. 
Already, complex ecological simulators -- SORTIE in forest ecosystems, EcoPath with EcoSim in marine systems, GCM models in climate -- are the Atari games of our policy grandmasters.
Detailed, process-based models allow a window into the future scenarios of ecological management.
Even simple processed-based ecological models can provide important insight into future scenarios that no amount of historical data would reveal.
Rather than displacing existing ecological models and knowledge, we illustrate how RL-based approaches can allow us to leverage such understanding to better explore and guide complex decision-making in changing environments.
Whereas previous decision techniques have historically constrained the complexity and realism of practical modeling efforts, RL may open the door to the design of ever more realistic assumptions. 
Drawing on examples from fisheries management, ecosystem tipping points, and wildfire spread, we illustrate how ecological models and simulations can be used to successfully train RL agents that match or exceed the performance of available management strategies.
Although our examples are primarily a proof-of-principle, they serve to illuminate both the promise as well as potential pitfalls.
